---
title: "GLMM Maternal Mortality Federally- Summer 2025"
subtitle: "This is a Slide Template"
author: "Carolyn Herrera & Catherine Funte (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction  {.smaller}
Generalized Linear Mixed Models (GLMMs) are a flexible class of statistical models that combine the features of two powerful tools: Generalized Linear Models (GLMs) and Mixed-Effects Models[@agresti2015foundations]. Like GLMs, GLMMs can model non-normal outcome variables, such as binary, count, or proportion data. However, they go a step further by incorporating random effects, which account for variation due to grouping or clustering in the data, correlated observations, and overdispersion.

In practical terms, GLMMs are especially useful when data points are not independent, such as when students are nested within schools, patients are treated within hospitals, or repeated measures are taken from the same subject over time. For example, Thall wrote that issues with longitudinal clinical trial basic count data from repeated measures taken from the same subject over time will have problems detecting comparable between subject outcomes because it can be difficult to determine if outcomes are time dependent or due to treatment groups, thus a general linear mixed model method may be utilized to represent dependence upon each patient, incorporate covariate data, create time as a function, account for variability between patients,and  be flexible and tractable [@thall1988mixed]. The random effects help model the correlation within clusters and allow for unobserved heterogeneity—differences that are not captured by the measured covariates.

GLMMs are good for:

-  Handling hierarchical or grouped data (e.g., students within classrooms, patients within clinics)


-  Modeling non-normal outcomes, such as:


   -  Binary outcomes (using logistic GLMMs)


   -  Count data (using Poisson or negative binomial GLMMs)


   -  Proportions or rates


-  Improving inference by accounting for both fixed effects (predictors of interest) and random effects (random variation across groups)


-  Reducing bias and inflated Type I error rates that can result from ignoring data structure


GLMMs are ideal when your data is both complex in structure and involves non-Gaussian response variables, making them indispensable in fields like medicine, ecology, education, and social sciences. Tawiah et al describes zero-inflated Poisson GLMMs, an extension of Poisson GLMM that allows for overdispersion due to a prevalence of zeros in the data, common in health sector data[@tawiah2020zero]. The paper compares a Poisson GLM, a zero-inflated Poisson GLM, a Poisson GLMM, and a zero-inflated Poisson GLMM, applied to clustered maternal mortality data. Another paper by Owili et al utilizes a GLMM to investigate the impact of particulate matter on maternal and infant mortality globally [@owili2020impacts]. They use a Poisson link function and take year and country as random effects to account for differences in global data. 

We wish to analyze the data of federal maternal mortality deaths by utilizing a General linear mixed model with Poisson link as it is count data. We wish to see if ethnicity (a fixed effect) has any influence upon maternal death count by year(random effect). Like other public health or clinical data there will be issues such as correlated observations and overdispersion but GLMM will be utilized to parse through the noise and determine if indeed there are some patterns of maternal mortality among mothers of differing ethnicties.

In kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]\*

## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.



## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
